---
title: Expected lifetime
author: useR member
output:
    revealjs::revealjs_presentation:
        center: true
---


```{r loaddata, echo = FALSE}
set.seed(22)
# Read in data
# Each file has two columns.
# First column is species ID 1 == cutthroat, 2 == rainbow, 3 == grayling
# Second column is survival time in days since exposure.
# In obs.csv, survival time is time until death
# In cens.csv, survival time is time until end of experiment, or in otherwords,
# last known time alive
obs.data <- read.csv("./observed.csv", head = FALSE, col.names = c("species", "time"))
cens.data <- read.csv("./censored.csv", head = FALSE, col.names = c("species", "time"))

n1obs <- sum(obs.data$species == 1)
n2obs <- sum(obs.data$species == 2)
n3obs <- sum(obs.data$species == 3)
n1cens <- sum(cens.data$species == 1)
n2cens <- sum(cens.data$species == 2)
n3cens <- sum(cens.data$species == 3)
n1 <- n1obs + n1cens
n2 <- n2obs + n2cens
n3 <- n3obs + n3cens
count.data <- data.frame(species = c("cutthroat", "rainbow", "grayling"),
                         observed = c(n1obs, n2obs, n3obs),
                         censored = c(n1cens, n2cens, n3cens),
                         total = c(n1, n2, n3))
```

Setup
=====

How to use revealjs
-------------------

- First, install the revealjs package `install.packages("revealjs")`
- Now set the output in your yaml:

```
---
title: my title
author: my name
output:
    revealjs::revealjs_presentation
---
```

Build
-----

- Now build like normal!
- Click rstudio button or use `rmarkdown::render("file.Rmd")`
- See github readme for options:

[rstudio/revealjs](http://github.com/rstudio/revealjs)

Intro
=====

The study
---------

- Three species of fish, cutthroat, rainbow, and grayling exposed to biological
  agent.
- Time until death after exposure is recorded.
- Some observations are censored; we have last known time alive.

Objective
---------

- Estimate average time until death after exposure for each species.
- Difference in average time until death between cutthroat and rainbow.

Data
====

Number of fish
--------------

```{r number, echo = FALSE}
library(rbokeh)
ly_bar(figure(), species, total, data = count.data,
       hover = c(observed, censored))
```

Time until death
----------------

```{r timeuntildeath, echo = FALSE}
plot(0, 0, type = 'n', xlim = c(0, 4), ylim = c(0, 50), xaxt = 'n', xlab = '', ylab = '')
axis(1, at = c(1, 2, 3), label = c("cutthroat", "rainbow", "grayling"))
points(obs.data$time ~ jitter(obs.data$species), pch = 4)
cens.x <- jitter(cens.data$species)
points(cens.data$time ~ cens.x, pch = 1, cex = 2)
for (i in 1:nrow(cens.data))
{
    lines(c(cens.x[i], cens.x[i]), c(cens.data$time[i], 100), lty = 3, lwd = 1)
}
```

Model
=====

Parameters and variables
------------------------

- $\mu_j$ is the average lifetime of species $j$.
- $\sigma$ is the assumed constant across species standard deviation of
  lifetime.
- $y_i$ is the lifetime of the $i^\text{th}$ fish.
- $j[i]$ is the species of the $i^\text{th}$ fish.

Likelihood
----------

I assume lifetime follows a normal distribution.

$y_i \sim N(\mu_{j[i]}, \sigma^2)$


Prior
-----

$\mu_j \sim N(20, 10)$

$\sigma \sim N(0, 10)I_{\sigma>\infty}$

($\sigma$ is constrained to be positive)

Censored?
---------

- Likelihood tells us how we get our data from the parameters.
- We have $y\sim N(\mu, \sigma^2)$ and we want the probability of $y>C$.
- So, the likelihood contribution is the part of $N(\mu, \sigma^2)$ greater than $C$.

```{r chunkname, fig.keep = "last", fig.width = 3, fig.height = 3, echo = FALSE}
curve(dnorm(x, 30, 5), xlim = c(0, 50))
abline(v = 36)
```

Stan code
---------

```{r results = "asis", echo = FALSE}
cat("\`\`\`c++\n")
cat(readLines("./lifetime.stan"), sep = "\n")
cat("\`\`\`")
```

Fit model and evaluate
======================

```{r prepareforstan, echo = FALSE}
stan.data <- list(n1obs, n2obs, n3obs, n1cens, n2cens, n3cens,
                  y1obs = obs.data$time[obs.data$species == 1],
                  y2obs = obs.data$time[obs.data$species == 2],
                  y3obs = obs.data$time[obs.data$species == 3],
                  y1cens = cens.data$time[cens.data$species == 1],
                  y2cens = cens.data$time[cens.data$species == 2],
                  y3cens = cens.data$time[cens.data$species == 3])
```

```{r sample, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
library(rstan)
# samps <- stan(file = "./lifetime.stan", data = stan.data, chains = 4, iter = 4000)

# OR just load the data
load("./lifetime.RData")
```

Traceplot
---------

```{r traceplot, echo = FALSE}
plot(samps, plotfun = "trace", pars = c("mu1", "mu2", "mu3", "sigma"))
```

Summary
-------

```{r summary, echo = FALSE, results = "markup"}
library(pander)
pander(summary(samps)$summary[1:5, c(1, 3, 4, 6, 8)])
```

Results
=======

Average lifetime for cutthroat
------------------------------


```{r mu1hist, echo = FALSE}
library(htmltools)
library(htmlwidgets)
library(metricsgraphics)
library(RColorBrewer)

mu1 <- extract(samps, pars = "mu1")$mu1
mu2 <- extract(samps, pars = "mu2")$mu2

mjs_plot(as.numeric(mu1), width = 500, height = 500) %>%
mjs_histogram() %>%
mjs_labs(x_label = "Mean lifetime of cutthroat")

# mjs_add_marker(quantile(mu1, 0.025), "0.025 Quantile")
# mjs_add_marker(x_value = quantile(mu1, 0.975), label = "0.975 Quantile") %>%
# mjs_add_marker(x_value = mean(mu1), label = "mean") %>%
```

---


```{r mu1, echo = FALSE, fig.keep = "last"}
mu1 <- extract(samps, pars = "mu1")$mu1
hist(mu1)
abline(v = mean(mu1), lwd = 4)
abline(v = quantile(mu1, c(0.025, 0.975)), lwd = 2)
```

Difference in average lifetime between cutthroat and rainbow
------------------------------------------------------------

```{r diff, echo = FALSE}

mu1mu2 <- extract(samps, pars = "cut_rain")$cut_rain
mjs_plot(as.numeric(mu1mu2), width = 500, height = 500) %>%
mjs_histogram() %>%
mjs_labs(x_label = "Diff Mean lifetime")

# mjs_add_marker(quantile(mu1, 0.025), "0.025 Quantile")
# mjs_add_marker(x_value = quantile(mu1, 0.975), label = "0.975 Quantile") %>%
# mjs_add_marker(x_value = mean(mu1), label = "mean") %>%
```

---

```{r mu1_minus_mu2, echo = FALSE, fig.keep = "last"}
mu1mu2 <- extract(samps, pars = "cut_rain")$cut_rain
hist(mu1mu2)
abline(v = mean(mu1mu2), lwd = 4)
abline(v = quantile(mu1mu2, c(0.025, 0.975)), lwd = 2)
```

